IIT CS595: Topics \& Applications in Programming Languages\\ {\large Homework 2: Memory and Cost Semantics}

### 2021-11-06 HW2/hw2.tex
finish hw2

\title{IIT CS595: Topics \& Applications in Programming Languages\\
  {\large Homework 2: Memory and Cost Semantics}}
\author{Your Name Here}
\date{Out: Tuesday, Oct. 19\\
  Due: Wednesday, Nov. 3 11:59pm CDT\\
  %\vspace{0.5cm}
  %\updated{Sep. 15}
}

@@

{
  \centering
  \def \MathparLineskip {\lineskip=0.43cm}
  \begin{mathpar}
    \Rule{E-1}
         {v\val}
         {\eval[0]{v}{v}}
    \and
    \Rule{E-2}
         {\eval[\blank]{e_1}{\kwfun{x}{\tau}{e}}\\
           \eval[\blank]{e_2}{v}\\
           \eval[\blank]{\sub{v}{x}{e}}{v'}}
         {\eval[\blank]{\kwapp{e_1}{e_2}}{v'}}
    \and
    \Rule{E-3}
         {\eval[\blank]{e_1}{v_1}\\
           \eval[\blank]{e_2}{v_2}
         }
         {\eval[\blank]{\kwpair{e_1}{e_2}}{\kwvpair{v_1}{v_2}}}
    \and
    \Rule{E-4}
         {\eval[\blank]{e}{\kwvpair{v_1}{v_2}}}
         {\eval[\blank]{\kwfst{e}}{v_1}}
    \and
    \Rule{E-5}
         {\eval[\blank]{e}{\kwvpair{v_1}{v_2}}}
         {\eval[\blank]{\kwsnd{e}}{v_2}}
  \end{mathpar}
}

@@

\begin{proof}
  Your proof here.
\end{proof}

@@

\begin{proof}
  Your proof here.
\end{proof}

@@

  \item (E-2).
    Then~$e = \kwapp{e_1}{e_2}$
    and~$\eval[n_1]{e_1}{\kwfun{x}{\tau}{e}}$
    and~$\eval[n_2]{e_2}{v}$
    and~$\eval[n_3]{\sub{v}{x}{e}}{v'}$ and~$n = n_1 + n_2 + n_3$.
    By induction,
    \[
    \cfg{\store}{e_1} \step^*
    \cfg{\stjoin{\store}{\store_1}}{\kwfun{x}{\tau}{e}}
    \text{~and~}\cfg{\stjoin{\store}{\store_1}}{e_2} \step^*
    \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v}
    \text{~and~}\cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}
    {\sub{v}{x}{e}} \step^*
    \cfg{\stjoin{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{\store_3}}
        {v'}
    \]
    and
    \[|\store_1| = n_1\text{~and~}|\store_2| = n_2\text{~and~}|\store_3| = n_3\]
    By Lemma~\ref{lem:compose-exps},
    $\cfg{\store}{e} \step^*
    \cfg{\stjoin{\store}{\store_1}}{\kwapp{(\kwfun{x}{\tau}{e})}{e_2}}$
    and~$\cfg{\stjoin{\store}{\store_1}}{\kwapp{(\kwfun{x}{\tau}{e})}{e_2}}
    \step^*
    \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}
          {\kwapp{(\kwfun{x}{\tau}{e})}{v}}$.\\
          By Lemma~\ref{lem:compose-steps},
      $\cfg{\store}{e} \step^*
        \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}
            {(\kwapp{\kwfun{x}{\tau}{e})}{v}}$.\\
      By (S-3) and Lemma~\ref{lem:compose-steps},
      $\cfg{\store}{e} \step^*
      \cfg{\stjoin{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{\store_3}}
          {v'}$.
          We have
      $|\stjoin{\store_1}{\stjoin{\store_2}{\stsing{\kwvpair{v_1}{v_2}}}}| =
      |\store_1| + |\store_2| + |\store_3| = n$.

\item (E-1).
\item (E-3).
\item (E-4).
\item (E-5).
  \end{itemize}
\end{proof}

@@

\begin{lemma}\label{lem:closed}
  If~$\eval[n]{e'}{v}$
  and~$\cfg{\store}{e} \step \cfg{\stjoin{\store}{\store'}}{e'}$
  then~$\eval[n+k]{e}{v}$
  and~$|\store'| = k$.
\end{lemma}
  \begin{proof}
    By induction on the derivation of 
    $\cfg{\store}{e} \step \cfg{\stjoin{\store}{\store'}}{e'}$.
    \begin{itemize}
    \item (S-1)
      Then~$e = \kwapp{e_1}{e_2}$ and~$e' = \kwapp{e_1'}{e_2}$.
      By inversion on (E-2),
      $\eval[n_1]{e_1'}{\kwfun{x}{\tau}{e}}$
      and~$\eval[n_2]{e_2}{v}$
      and~$\eval[n_3]{\sub{v}{x}{e}}{v'}$
      and~$n = n_1 + n_2 + n_3$.
      By induction,
      $\eval[n_1 + k]{e_1}{\kwfun{x}{\tau}{e}}$
      and~$|\store'| = k$.
      By (E-2),
      $\eval[n+k]{e}{v}$.

    \item Your other cases here...
    \end{itemize}
  \end{proof}

>>>>

\title{IIT CS595: Topics \& Applications in Programming Languages\\
  {\large Homework 2: Memory and Cost Semantics}}
\author{Dustin Van Tate Testa}
\date{Out: Tuesday, Oct. 19\\
  Due: Wednesday, Nov. 3 11:59pm CDT\\
  %\vspace{0.5cm}
  %\updated{Sep. 15}
}

@@

{
  \centering
  \def \MathparLineskip {\lineskip=0.43cm}
  \begin{mathpar}
    \Rule{E-1}
         {v\val}
         {\eval[0]{v}{v}}
    \and
    \Rule{E-2}
         {\eval[n_1]{e_1}{\kwfun{x}{\tau}{e}}\\
           \eval[n_2]{e_2}{v}\\
           \eval[n_3]{\sub{v}{x}{e}}{v'}}
         {\eval[n_1 + n_2 + n_3]{\kwapp{e_1}{e_2}}{v'}}
    \and
    \Rule{E-3}
         {\eval[n_1]{e_1}{v_1}\\
           \eval[n_2]{e_2}{v_2}
         }
         {\eval[n_1 + n_2 + 1]{\kwpair{e_1}{e_2}}{\kwvpair{v_1}{v_2}}}
    \and
    \Rule{E-4}
         {\eval[n]{e}{\kwvpair{v_1}{v_2}}}
         {\eval[n]{\kwfst{e}}{v_1}}
    \and
    \Rule{E-5}
         {\eval[n]{e}{\kwvpair{v_1}{v_2}}}
         {\eval[n]{\kwsnd{e}}{v_2}}
  \end{mathpar}
}

@@

\begin{proof}
  By induction on the derivation of
    $\cfg{\store}{e} \step^* \cfg{\store'}{e'}$.
    \begin{itemize}
    \item (M-1). Then~$e = e'$ and~$\store = \store'$.
      By (M-1),
      $\cfg{\store}{\kwfst{e}} \step^*
      \cfg{\store}{\kwfst{e}}$.
    \item (M-2)
      Then~$\cfg{\store}{e} \step \cfg{\store''}{e''}$ 
      and $\cfg{\store''}{e''} \step^* \cfg{\store}{e'}$.\\
      By (S-7), 
      $\cfg{\store}{\kwfst{e}} \step \cfg{\store''}{\kwfst{e''}}$.
      By induction,
      $\cfg{\store''}{\kwfst{e''}} \step^* \cfg{\store'}{\kwfst{e'}}$
      so by (M-2),
      $\cfg{\store}{\kwfst{e}} \step^* \cfg{\store'}{\kwfst{e'}}$
    \end{itemize}
\end{proof}

@@

\begin{proof}
  Your proof here.
  \begin{itemize}
    \item by induction on derivation of the step judgement, either
        \begin{itemize}
            \item $e = e' = e''$ and $\store = \store' = \store''$:\\
                in which case for both step judgements (M-1) would be applied

            \item $e' = e''$ and $\store' = \store''$:\\
                in which case (M-2) is applied to the first step and (M-1) applied after
                % \begin{mathpar}
                %     \inferrule*
                %     {\inferrule* {$ $}{\cfg{\store}{e} \step \cfg{\store'''}{e'''}}\\
                %       \inferrule* {$ $}{\cfg{\store'''}{e'''} \step^* \cfg{\store'}{e'}}}
                %     {\cfg{\store}{e} \step^* \cfg{\store'}{e'}}
                % \end{mathpar}
            \item $e = e'$ and $\store = \store'$:\\
                in which case (M-1) is applied to make the first step and then (M-2) after

            \item all $e$ and all $\store$ values are unique:\\
                in which case (M-2) is applied twice
        \end{itemize}
  \end{itemize}
\end{proof}

@@

  \item (E-2).
    Then~$e = \kwapp{e_1}{e_2}$
    and~$\eval[n_1]{e_1}{\kwfun{x}{\tau}{e}}$
    and~$\eval[n_2]{e_2}{v}$
    and~$\eval[n_3]{\sub{v}{x}{e}}{v'}$ and~$n = n_1 + n_2 + n_3$.
    By induction,
    \[
    \cfg{\store}{e_1} \step^*
    \cfg{\stjoin{\store}{\store_1}}{\kwfun{x}{\tau}{e}}
    \text{~and~}\cfg{\stjoin{\store}{\store_1}}{e_2} \step^*
    \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v}
    \text{~and~}\cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}
    {\sub{v}{x}{e}} \step^*
    \cfg{\stjoin{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{\store_3}}
        {v'}
    \]
    and
    \[|\store_1| = n_1\text{~and~}|\store_2| = n_2\text{~and~}|\store_3| = n_3\]
    By Lemma~\ref{lem:compose-exps},
    $\cfg{\store}{e} \step^*
    \cfg{\stjoin{\store}{\store_1}}{\kwapp{(\kwfun{x}{\tau}{e})}{e_2}}$
    and~$\cfg{\stjoin{\store}{\store_1}}{\kwapp{(\kwfun{x}{\tau}{e})}{e_2}}
    \step^*
    \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}
          {\kwapp{(\kwfun{x}{\tau}{e})}{v}}$.\\
          By Lemma~\ref{lem:compose-steps},
      $\cfg{\store}{e} \step^*
        \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}
            {(\kwapp{\kwfun{x}{\tau}{e})}{v}}$.\\
      By (S-3) and Lemma~\ref{lem:compose-steps},
      $\cfg{\store}{e} \step^*
      \cfg{\stjoin{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{\store_3}}
          {v'}$.
          We have
      $|\stjoin{\store_1}{\stjoin{\store_2}{\store_3}}| =
      |\store_1| + |\store_2| + |\store_3| = n$.

\item (E-1). Then $e \val$ and $\eval[0]{e}{v}$ and $e = v$ and $n = 0$ which by lemma 1 translates to $\cfg{\estore}{e} \step^* \cfg{\store}{e}$. Because values cannot step according to the defined small-step semantics rules (exhaustion of rules (V- 1..3)), the $\step^*$ judgement must follow rule (M-1), thus $\estore = \store$ and $|\store| = n = 0$.

\item (E-3).
    Then~$e = \kwpair{e_1}{e_2}$
    and~$\eval[n_1]{e_1}{v_1}$
    and~$\eval[n_2]{e_2}{v_2}$
    and~$\eval[1]{\kwpair{v_1}{v_2}}{\kwvpair{v_1}{v_2}}$
    and~$n = n_1 + n_2 + 1$.
    By induction,
    \[
    \cfg{\store}{e_1} 
    \step^* \cfg{\stjoin{\store}{\store_1}}{v_1}
        \text{~and~}
    \cfg{\store}{e_2}
    \step^* \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_2}
        \text{~and~}
    \cfg{\store}{e_2} 
    \step^* \cfg{\stjoin{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{\store_3}}{v_2}
    \]
    and
    \[|\store_1| = n_1\text{~and~}|\store_2| = n_2\text{~and~}|\store_3| = 1\]
    By Lemma~\ref{lem:compose-exps},
    $
        \cfg{\store}{e} 
        \step^* \cfg{\stjoin{\store}{\store_1}}{\kwpair{v_1}{e_1}}
    $and~$
        \cfg{\stjoin{\store}{\store_1}}{\kwpair{v_1}{e_1}}
        \step^* \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{\kwpair{v_1}{v_2}}
    $.\\
    By Lemma~\ref{lem:compose-steps},
      $\cfg{\store}{e} 
        \step^*
         \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{\kwpair{v_1}{v_2}}$.\\
    By (S-6) and Lemma~\ref{lem:compose-steps},
      $\cfg{\store}{e} \step^*
      \cfg{\stjoin{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{\stsing{\kwvpair{v_1}{v_2}}}}
          {\kwvpair{v_1}{v_2}}$.
          We have
      $|\stjoin{\store_1}{\stjoin{\store_2}{\stsing{\kwvpair{v_1}{v_2}}}}| =
      |\store_1| + |\store_2| + |\stsing{\kwvpair{v_1}{v_2}}| = n$.

\item (E-4).
    Then~$e = \kwfst{e_1}$
    and~$\eval[n_1]{e_1}{\kwvpair{v_1}{v_2}}$
    and~$\eval[0]{e}{v_1}$
    and~$n = n_1 + 0$.
    By induction,
    \[
    \cfg{\store}{e_1} \step^* \cfg{\stjoin{\store}{\store_1}}{\kwvpair{v_1}{v_2}}
    \text{~and~}
    \cfg{\stjoin{\store}{\store_1}}{\kwfst{\kwvpair{v_1}{v_2}}} 
        \step^* 
        \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_1}
    \]
    and
    \[|\store_1| = |\store_2| = n_1 \text{~and~} n = n_1 \]
    By Lemma~\ref{lem:compose-exps},
    $
        \cfg{\store}{e} \step^* \cfg{\stjoin{\store}{\store_1}}{\kwfst{\kwvpair{v_1}{v_2}}}
    $~and~$
        \cfg{\stjoin{\store}{\store_1}}{\kwfst{\kwvpair{v_1}{v_2}}} 
        \step^* 
        \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_1}
    $.\\
    By Lemma~\ref{lem:compose-steps},
      $\cfg{\store}{e} \step^*
        \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_1}$.\\
    By (S-8) and Lemma~\ref{lem:compose-steps},
      $\cfg{\store}{e} \step^*
      \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_1}$.
          We have
      $|\stjoin{\store_1}{\store_2}| =
      |\store_1| = n$.

\item (E-5).
    Then~$e = \kwsnd{e_1}$
    and~$\eval[n_1]{e_1}{\kwvpair{v_1}{v_2}}$
    and~$\eval[0]{e}{v_1}$
    and~$n = n_1 + 0$.
    By induction,
    \[
    \cfg{\store}{e_1} \step^* \cfg{\stjoin{\store}{\store_1}}{\kwvpair{v_1}{v_2}}
    \text{~and~}
    \cfg{\stjoin{\store}{\store_1}}{\kwsnd{\kwvpair{v_1}{v_2}}} 
        \step^* 
        \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_2}
    \]
    and
    \[|\store_1| = |\store_2| = n_1 \text{~and~} n = n_1 \]
    By Lemma~\ref{lem:compose-exps},
    $
        \cfg{\store}{e} \step^* \cfg{\stjoin{\store}{\store_1}}{\kwsnd{\kwvpair{v_1}{v_2}}}
    $~and~$
        \cfg{\stjoin{\store}{\store_1}}{\kwsnd{\kwvpair{v_1}{v_2}}} 
        \step^* 
        \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_2}
    $.\\
    By Lemma~\ref{lem:compose-steps},
      $\cfg{\store}{e} \step^*
        \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_2}$.\\
    By (S-9) and Lemma~\ref{lem:compose-steps},
      $\cfg{\store}{e} \step^*
      \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_2}$.
          We have
      $|\stjoin{\store_1}{\store_2}| =
      |\store_1| = n$.

  \end{itemize}
\end{proof}

@@

\begin{lemma}\label{lem:closed}
  If~$\eval[n]{e'}{v}$
  and~$\cfg{\store}{e} \step \cfg{\stjoin{\store}{\store'}}{e'}$
  then~$\eval[n+k]{e}{v}$
  and~$|\store'| = k$.
\end{lemma}
  \begin{proof}
    By induction on the derivation of 
    $\cfg{\store}{e} \step \cfg{\stjoin{\store}{\store'}}{e'}$.
    \begin{itemize}
    \item (S-1)
      Then~$e = \kwapp{e_1}{e_2}$ and~$e' = \kwapp{e_1'}{e_2}$.
      By inversion on (E-2),
      $\eval[n_1]{e_1'}{\kwfun{x}{\tau}{e}}$
      and~$\eval[n_2]{e_2}{v}$
      and~$\eval[n_3]{\sub{v}{x}{e}}{v'}$
      and~$n = n_1 + n_2 + n_3$.
      By induction,
      $\eval[n_1 + k]{e_1}{\kwfun{x}{\tau}{e}}$
      and~$|\store'| = k$.
      By (E-2),
      $\eval[n+k]{e}{v'}$.

    \item (S-2) Then~$e = \kwapp{\kwfun{x}{\tau}{e_1}}{e_2}$
        and~$e' = \kwapp{\kwfun{x}{\tau}{e_1}}{e_2'}$.
        By inversion on (E-2),
        $\eval[0]{\kwfun{x}{\tau}{e_1}}{\kwfun{x}{\tau}{e_1}}$
        and~$\eval[n_2]{e_2}{v'}$
        and~$\eval[n_3]{\sub{v'}{x}{e_1}}{v}$
        and~$n = 0 + n_2 + n_3$.
        By induction,
        $\eval[n' + k]{e_2}{v'}$
        and~$|\store'| = k$.
        By (E-2),
        $\eval[n+k]{e}{v}$.

    \item (S-3) Then~$e = \kwapp{\kwfun{x}{\tau}{e_1}}{v'}$
        and~$e' = \kwapp{\kwfun{x}{\tau}{e_1}}{e_2'}$
        and~$v' \val$.
        By inversion on (E-2),
        $\eval[0]{\kwfun{x}{\tau}{e_1}}{\kwfun{x}{\tau}{e_1}}$
        and~$\eval[0]{v'}{v'}$
        and~$\eval[n_1]{\sub{v'}{x}{e_1}}{v}$
        and~$n = 0 + 0 + n_1$.
        By induction,
        $\eval[n' + k]{\sub{v'}{x}{e_1}}{v}$
        $|\store'| = k$.
        By (E-2),
        $\eval[n+k]{e}{v}$.

    \item (S-4) Then~$e = \kwpair{e_1}{e_2}$
        and~$e' = \kwpair{e_1'}{e_2}$.
        By inversion on (E-3),
        $\eval[n_1]{e_1}{v_1}$
        and~$\eval[n_2]{e_2}{v_2}$
        and~$\eval[1]{\kwpair{v_1}{v_2}}{\kwvpair{v_1}{v_2}}$
        and~$n = n_1 + n_2 + 1$.
        By induction,
        $\eval[n_1 + k]{e_1}{v_1}$
        $|\store'| = k$.
        By (E-3),
        $\eval[n+k]{e}{v}$
    \item (S-5) Then~$e = \kwpair{v_1}{e_2}$
        and~$e' = \kwpair{v_1}{e_2'}$.
        By inversion on (E-3),
        $\eval[0]{v_1}{v_1}$
        and~$\eval[n_2]{e_2}{v_2}$
        and~$\eval[1]{\kwpair{v_1}{v_2}}{\kwvpair{v_1}{v_2}}$
        and~$n = 0 + n_2 + 1$.
        By induction,
        $\eval[n_2 + k]{e_2}{v_2}$
        $|\store'| = k$.
        By (E-3),
        $\eval[n+k]{e}{v}$
    \item (S-6) Then~$e = \kwpair{v_1}{v_2}$
        and~$e' = \kwpair{v_1}{v_2}$.
        By inversion on (E-3),
        $\eval[0]{v_1}{v_1}$
        and~$\eval[0]{v_2}{v_2}$
        and~$\eval[1]{\kwpair{v_1}{v_2}}{\kwvpair{v_1}{v_2}}$
        and~$n = 0 + 0 + 1$.
        By induction,
        $\eval[n_2 + k]{e}{\kwvpair{v_1}{v_2}}$
        $|\store'| = k = 1$.
        By (E-3),
        $\eval[n+k]{e}{v}$

    \item (S-7) Then~$e = \kwfst{e_1}$ and~$e' = \kwfst{e_1'}$.
        By inversion on (E-4),
        $\eval[n_1]{e_1}{\kwvpair{v_1}{v_2}}$
        and~$\eval[n_1]{\kwfst{e_1}}{v_1}$
        and~$n = n_1$.
        By induction,
        $\eval[n_1 + k]{e}{v_1}$
        $|\store'| = k$.
        By (E-4),
        $\eval[n+k]{e}{v}$

    \item (S-8) Then~$e = \kwfst{\kwvpair{v_1}{v_2}}$ and~$e' = v_1$.
        By inversion on (E-4),
        $\eval[0]{\kwvpair{v_1}{v_2}}{\kwvpair{v_1}{v_2}}$
        $\eval[0]{e'}{v_1}$, so $n = 0$.
        By (S-8), $\store' = \estore$, so $k = 0$.
        By (E-4),
        $\eval[n+k]{e}{v}$

    \item (S-9) Then~$e = \kwsnd{e_1}$ and~$e' = \kwsnd{e_1'}$.
        By inversion on (E-4),
        $\eval[n_1]{e_1}{\kwvpair{v_1}{v_2}}$
        and~$\eval[n_1]{\kwfst{e_1}}{v_2}$
        and~$n = n_1$.
        By induction,
        $\eval[n_1 + k]{e}{v_1}$
        $|\store'| = k$.
        By (E-5),
        $\eval[n+k]{e}{v}$

    \item (S-10) Then~$e = \kwfst{\kwvpair{v_1}{v_2}}$ and~$e' = v_1$.
        By inversion on (E-5),
        $\eval[0]{\kwvpair{v_1}{v_2}}{\kwvpair{v_1}{v_2}}$
        $\eval[0]{e'}{v_2}$, so $n = 0$.
        By (S-10), $\store' = \estore$, so $k = 0$.
        By (E-5),
        $\eval[n+k]{e}{v}$

    \end{itemize}
  \end{proof}
